refnx - Neutron and X-ray reflectometry analysis in Python

### 2018-07-19 paper/manuscript.tex
PAPER: flesh out lipid example section

\newcommand{\refnx}{\emph{refnx}}
\newcommand{\Objective}{\textbf{Objective}}
\newcommand{\GlobalObjective}{\textbf{GlobalObjective}}
\newcommand{\Parameter}{\textbf{Parameter}}
\newcommand{\Structure}{\textbf{Structure}}
\newcommand{\Slab}{\textbf{Slab}}
\newcommand{\Component}{\textbf{Component}}
\newcommand{\LipidLeaflet}{\textbf{LipidLeaflet}}
\newcommand{\Transform}{\textbf{Transform}}
\newcommand{\DataD}{\textbf{Data1D}}
\newcommand{\ReflectModel}{\textbf{ReflectModel}}
\newcommand{\Spline}{\textbf{Spline}}
\newcommand{\conda}{\emph{conda}}
\newcommand{\corner}{\emph{corner}}
\newcommand{\MixedReflectModel}{\emph{MixedReflectModel}}
\newcommand{\pip}{\emph{pip}}
\newcommand{\emcee}{\emph{emcee}}
\newcommand{\NumPy}{\emph{NumPy}}
\newcommand{\SciPy}{\emph{SciPy}}
\newcommand{\Cython}{\emph{Cython}}
\newcommand{\Jupyter}{\emph{Jupyter}}
\newcommand{\ipywidgets}{\emph{ipywidgets}}

@@

The whole package is designed with a focus on reproducible research; its use in \Jupyter\ notebooks (which includes a graphical user interface), and subsequent distribution of those notebooks as supplementary information, permits scientists to straightforwardly reproduce analyses made by others.
\end{abstract}

@@

Given that there are more and more publications arising from the reflectometry technique, it is vital that the analysis is reproducible. Reproducibility in research is an underlying principle of science. Unfortunately, it is not always possible to reproduce the results of others \cite{Stark2018}, because there is frequently not enough information provided in journal articles to repeat the analyses. Even if the datasets and software packages used to analyse them are supplied in supplementary information (most often they are not), a comprehensive, ordered, set of instructions would need to be provided.
One example for addressing this reproducibility issue are the guidelines in the small-angle scattering community for the deposition of data and and associated models \cite{Trewhella:jc5010}.

Here we outline a new reflectometry analysis package, \refnx\ (version number 0.1 is used in this paper \cite{refnx}), that can address the reproducibility issue for the reflectometry community as it is specifically designed for use in \Jupyter\ notebooks \cite{Kluyver:2016aa} using a Python kernel \footnote{We do not mean that other programs are irreproducible, rather that the information provided by scientists in journal articles is often lacking.}.
These notebooks are a mix of executable code cells and markup. An analysis performed in such a notebook can be appended as supplementary information, along with the data, to an article. Readers can then replicate the exact data analysis, provided they have set up the same computing environment. Setting up the computing environment is simplified using the \conda\ package manager \cite{conda}, and an environment file (although other approaches are available). To that end the \Jupyter\ notebooks for the example analysis performed in this paper are appended as supplementary information.

@@

The \Structure\ object represents the interfacial model, comprising of individual \Component\ objects in series. Each \Component\ represents a subset of the interface and contains several physically relevant Parameters. The simplest \Component\ is a \Slab\, which has a uniform scattering length density (SLD), thickness, roughness, and volume fraction of solvent. The simplest models are simply a series of \Slab\. More sophisticated components include \LipidLeaflet\ and \Spline\. \Spline\ is designed for freeform modelling of an SLD profile using spline interpolation. In addition to its constituent parameters each \Component\ can also contribute to the log-prior probability. This is useful when a \Component\ has a derived value, such as surface coverage, which is already known, and can be included as prior knowledge of the system. Each \Component\ has a \emph{slabs} property that represents the strata for its particular subset of the interface. A \Slab\ object has a single slice because it is a single thickness of uniform SLD. A \LipidLeaflet\ can be made of two slices (head/tail regions), but the \Spline\ Component has many. These slices have uniform SLD, and use the Nevot-Croce approach \cite{Nevot1980} to describe roughness between them. If a different roughness model is required then a 'microsliced' \Component\ could be written to accomodate that.

@@

\subsection*{Example}
Neutron reflectometry is an ideal technique for the study of biologically relevant lipid membrane mimics and their interactions with proteins, etc. Multiple contrast variation measurements are necessary to reduce modelling ambiguity (due to loss of phase information in the scattering experiment) and improve the ability to determine the structure of various components in the system. The gold standard approach for analysis of these datasets is co-refinement with a common model, and to parameterise the model in terms of chemically relevant parameters, such as area per molecule. Sometimes a patchy coverage (distinct to low area per molecule) necessitates the use of an (incoherent) sum of reflectivities from different areas. \refnx\ has functionality for all these requirements, such as the \LipidLeaflet\ component for describing the head and tail groups of a lipid leaflet, and \MixedReflectModel\. It is straightforward to develop/modify new components for different structural functionality, a consequence of the object-oriented approach. 
%describe lipidLeaflet TODO The \LipidLeaflet\ component
Here \LipidLeaflet\ is used to co-refine three contrasts (\ce{D_2O}, \ce{Si} contrast match [hdmix, SLD=$2.07\times10^{-6}\AA^{-2}$], and \ce{H_2O}) of a 1,2-dimyristoyl-sn-glycero-3-phosphocholine (DMPC) bilayer at the solid-liquid interface, Figure \ref{fig:global_fit} \footnote{The validity of \LipidLeaflet\ does depend on the area per molecule being equal for the headgroup and tailgroup regions, as pointed out by Gerelli \cite{Gerelli2016}, which can be violated if there are guest molecules that insert in the membrane.}. The \Jupyter\ notebook used for the analysis, \emph{lipid.ipynb}, is available in the supplementary information. The corner plot (Figure \ref{fig:corner}) produced from the MCMC analysis shows the covariance between parameters, with an area per molecule of 57.0$\AA\pm0.15$. Figures~\ref{fig:global_fit} and \ref{fig:d2o_sld_spread} show the probability distribution of the generative model around the data and in the SLD profile. These are obtained by random samples from the MCMC chain. The spread in SLD profiles is used to determine what range of structures is consistent with the data. Multi-modalities in these SLD profiles can be due to statistical uncertainties, the Q ranges measured, and the loss of phase information in NR \cite{Majkrzak1999, Heinrich2009}.
%TODO use exact area per molecule, and use a reference for the Xray analysis

@@

\section*{Comments on reproducibility of analyses}
In order for a given scattering analysis to be fully reproducible by others a general set of conditions need to be met:
\begin{itemize}
  \item the datasets used need to be deposited with a journal article, or be freely available.
  \item the exact software environment needs to be recreatable.
  \item the exact ordered set of steps taken during the analysis needs to be listed.
\end{itemize} 
Each of these points are often omitted in the literature. For example, the use of different software versions may change the output of an analysis, or the use of a GUI program may not record the full set of steps, or options, applied by a user.
Whilst it is unable to meet the first criteria by itself, the use of \refnx\ in a \Jupyter\ notebook can fulfil the other two requirements, providing a little care is taken. The exact software environment can be recreated by noting down the versions of the software packages used during an analysis (\refnx\, \SciPy\, \NumPy\, Python, etc). At a later date those exact versions can be installed in the same Python version using one of: the \conda\ package manager, by installing from the source at a given version tag in the git repository, or by \pip\. \conda\ can use environment files to recreate a specific setup. An alternative way of recreating the environment is by using a virtual machine, such as Docker. The ordered set of steps to perform the analysis is recordable in the \Jupyter\ notebook.
The usefulness of open-source software in a git (or other version controlled) repository must be emphasised here. With closed source or proprietary software the ability to return to a specific software version/environment can be difficult, different versions can have modifications that can change the output of an analysis. In addition accessibility (due to cost, etc) to the wider scientific community can also hinder reproducibility.
Moreover, there are important ramifications for verifiability. \refnx\ is based on a fully open software stack, with good unit test coverage. The user can run tests for each component and inspect parts for correctness. For example, the behaviour of the reflectivity calculation in \refnx\ is checked from first principles in the test suite; and can be done now and in several years time. If problems are discovered, they can be corrected. With a closed source program such checking is harder, one has to examine the inputs/outputs of a black box without knowing what is happening in between, making it harder to find errors; one has to rely on trust. Whilst this may not be an issue for relatively simple calculations, it may be for complex algorithms, such as those involved with minimisation.

@@

\section*{Acknowledgements}
We acknowledge Anton Le Brun (ANSTO) for the provision of the lipid bilayer datasets in the example, James Hester (ANSTO) for comments made on the draft manuscript, and Andrew McCluskey (Bath University) and Isaac Gresham (UNSW) for important feedback on refnx development.

>>>>

\newcommand{\refnx}{\emph{refnx}}
\newcommand{\Objective}{\textbf{Objective}}
\newcommand{\GlobalObjective}{\textbf{GlobalObjective}}
\newcommand{\Parameter}{\textbf{Parameter}}
\newcommand{\Structure}{\textbf{Structure}}
\newcommand{\Slab}{\textbf{Slab}}
\newcommand{\Component}{\textbf{Component}}
\newcommand{\LipidLeaflet}{\textbf{LipidLeaflet}}
\newcommand{\Transform}{\textbf{Transform}}
\newcommand{\DataD}{\textbf{Data1D}}
\newcommand{\ReflectModel}{\textbf{ReflectModel}}
\newcommand{\Spline}{\textbf{Spline}}
\newcommand{\conda}{\emph{conda}}
\newcommand{\corner}{\emph{corner}}
\newcommand{\MixedReflectModel}{\textbf{MixedReflectModel}}
\newcommand{\pip}{\emph{pip}}
\newcommand{\emcee}{\emph{emcee}}
\newcommand{\NumPy}{\emph{NumPy}}
\newcommand{\SciPy}{\emph{SciPy}}
\newcommand{\Cython}{\emph{Cython}}
\newcommand{\Jupyter}{\emph{Jupyter}}
\newcommand{\ipywidgets}{\emph{ipywidgets}}

@@

The whole package is designed with a focus on reproducible research; its use in \Jupyter\ notebooks (which includes a graphical user interface), and subsequent distribution of those notebooks as supporting information, permits scientists to straightforwardly reproduce analyses made by others.
\end{abstract}

@@

Given that there are more and more publications arising from the reflectometry technique, it is vital that the analysis is reproducible. Reproducibility in research is an underlying principle of science. Unfortunately, it is not always possible to reproduce the results of others \cite{Stark2018}, because there is frequently not enough information provided in journal articles to repeat the analyses. Even if the datasets and software packages used to analyse them are supplied in supporting information (most often they are not), a comprehensive, ordered, set of instructions would need to be provided.
One example for addressing this reproducibility issue are the guidelines in the small-angle scattering community for the deposition of data and and associated models \cite{Trewhella:jc5010}.

Here we outline a new reflectometry analysis package, \refnx\ (version number 0.1 is used in this paper \cite{refnx}), that can address the reproducibility issue for the reflectometry community as it is specifically designed for use in \Jupyter\ notebooks \cite{Kluyver:2016aa} using a Python kernel \footnote{We do not mean that other programs are irreproducible, rather that the information provided by scientists in journal articles is often lacking.}.
These notebooks are a mix of executable code cells and markup. An analysis performed in such a notebook can be appended as supporting information, along with the data, to an article. Readers can then replicate the exact data analysis, provided they have set up the same computing environment. Setting up the computing environment is simplified using the \conda\ package manager \cite{conda}, and an environment file (although other approaches are available). To that end the \Jupyter\ notebooks for the example analysis performed in this paper are appended as supporting information.

@@

The \Structure\ object represents the interfacial model, comprising of individual \Component\ objects in series. Each \Component\ represents a subset of the interface and contains several physically relevant Parameters. The simplest \Component\ is a \Slab\, which has a uniform scattering length density (SLD), thickness, roughness, and volume fraction of solvent. The simplest models are simply a series of \Slab. More sophisticated components include \LipidLeaflet\ and \Spline. \Spline\ is designed for freeform modelling of an SLD profile using spline interpolation. In addition to its constituent parameters each \Component\ can also contribute to the log-prior probability. This is useful when a \Component\ has a derived value, such as surface coverage, which is already known, and can be included as prior knowledge of the system. Each \Component\ has a \emph{slabs} property that represents the strata for its particular subset of the interface. A \Slab\ object has a single slice because it is a single thickness of uniform SLD. A \LipidLeaflet\ can be made of two slices (head/tail regions), but the \Spline\ Component has many. These slices have uniform SLD, and use the Nevot-Croce approach \cite{Nevot1980} to describe roughness between them. If a different roughness model is required then a 'microsliced' \Component\ could be written to accomodate that.

@@

\subsection*{Example}
Neutron reflectometry is an ideal technique for the study of biologically relevant lipid membrane mimics and their interactions with proteins, etc. Multiple contrast variation measurements are necessary to reduce modelling ambiguity (due to loss of phase information in the scattering experiment) and improve the ability to determine the structure of various components in the system. The gold standard approach for analysis of these datasets is co-refinement with a common model, and to parameterise the model in terms of chemically relevant parameters, such as area per molecule. Sometimes a patchy coverage (distinct to low area per molecule) necessitates the use of an (incoherent) sum of reflectivities from different areas. \refnx\ has functionality for all these requirements, such as the \LipidLeaflet\ component for describing the head and tail groups of a lipid leaflet, and \MixedReflectModel. It is straightforward to develop/modify new components for different structural functionality, a consequence of the object-oriented approach.

The parameters used in the \LipidLeaflet\ component are: area per molecule ($A$), thicknesses for each of the head and tail regions ($t_x$), sums of scattering lengths of the head and tail regions ($b_x$), partial volumes of the head and tail groups ($V_x$), roughness between head and tail region, and SLDs of the solvents for the head and tail group ($\rho_{x, solv}$). The overall SLD of each of the head and tail group regions are given by Equations~\ref{eqn:3} and \ref{eqn:4}. The approach used in this component ensures that there is a 1:1 correspondence of heads to tails. 
%TODO cite a paper which gives a whole load of volumes?
\begin{gather} 
\label{eqn:3} \phi_{x} = \frac{V_x}{At_x}\\
\rho_x =  \phi_{x} \frac{b_x}{V_x} + (1 - \phi_{x})\rho_{x, solv} \label{eqn:4}
\end{gather}

By default the head and tail solvents are assumed to be the same as the solvent that is used throughout the \Structure. This will be the case when using \LipidLeaflet\ for a solid-liquid reflectometry experiment. However, at the air-liquid, or liquid-liquid interfaces the solvent for the head and tail region may be different, and it is possible to use different solvent SLDs for each. We note that the \LipidLeaflet\ component may also be used to describe other amphiphiles adsorbing at an interface.

Here \LipidLeaflet\ is used to co-refine three contrasts (\ce{D_2O}, \ce{Si} contrast match [hdmix, SLD=$2.07\times10^{-6}\AA^{-2}$], and \ce{H_2O}) of a 1,2-dimyristoyl-sn-glycero-3-phosphocholine (DMPC) bilayer at the solid-liquid interface, Figure \ref{fig:global_fit} \footnote{The validity of \LipidLeaflet\ does depend on the area per molecule being equal for the headgroup and tailgroup regions, as pointed out by Gerelli \cite{Gerelli2016}, which can be violated if there are guest molecules that insert in the membrane.}. Two \LipidLeaflet\ are required to describe the inner and outer leaflets of a bilayer, the component contains an attribute which can reverse the direction of one of the leaflets. The use of individual objects to describe each leaflet leads to great flexibility; it becomes easy to model asymmetric bilayers (inner leaflet can be a different lipid to the outer lipid), and one can model interstitial water layers between the leaflets as well.

The \Jupyter\ notebook used for the analysis, \emph{lipid.ipynb}, is available in the supporting information. The corner plot (Figure \ref{fig:corner}) produced from the MCMC analysis shows the covariance between parameters, with an area per molecule of $57.0 \pm0.15 \AA$. Figures~\ref{fig:global_fit} and \ref{fig:d2o_sld_spread} show the probability distribution of the generative model around the data and in the SLD profile. These are obtained by random samples from the MCMC chain. The spread in SLD profiles is used to determine what range of structures is consistent with the data. Multi-modalities in these SLD profiles can be due to statistical uncertainties, the Q ranges measured, and the loss of phase information in NR \cite{Majkrzak1999, Heinrich2009}.
%TODO A reference for V_m?

@@

\section*{Comments on reproducibility of analyses}
In order for a given scattering analysis to be fully reproducible by others a general set of conditions need to be met:
\begin{itemize}
  \item the datasets used need to be deposited with a journal article, or be freely available.
  \item the exact software environment needs to be recreatable.
  \item the exact ordered set of steps taken during the analysis needs to be listed.
\end{itemize} 
Each of these points are often omitted in the literature. For example, the use of different software versions may change the output of an analysis, or the use of a GUI program may not record the full set of steps, or options, applied by a user.
Whilst it is unable to meet the first criteria by itself, the use of \refnx\ in a \Jupyter\ notebook can fulfil the other two requirements, providing a little care is taken. The exact software environment can be recreated by noting down the versions of the software packages used during an analysis (\refnx\, \SciPy\, \NumPy\, Python, etc). At a later date those exact versions can be installed in the same Python version using one of: the \conda\ package manager, by installing from the source at a given version tag in the git repository, or by \pip. \conda\ can use environment files to recreate a specific setup. An alternative way of recreating the environment is by using a virtual machine, such as Docker. The ordered set of steps to perform the analysis is recordable in the \Jupyter\ notebook.
The usefulness of open-source software in a git (or other version controlled) repository must be emphasised here. With closed source or proprietary software the ability to return to a specific software version/environment can be difficult, different versions can have modifications that can change the output of an analysis. In addition accessibility (due to cost, etc) to the wider scientific community can also hinder reproducibility.
Moreover, there are important ramifications for verifiability. \refnx\ is based on a fully open software stack, with good unit test coverage. The user can run tests for each component and inspect parts for correctness. For example, the behaviour of the reflectivity calculation in \refnx\ is checked from first principles in the test suite; and can be done now and in several years time. If problems are discovered, they can be corrected. With a closed source program such checking is harder, one has to examine the inputs/outputs of a black box without knowing what is happening in between, making it harder to find errors; one has to rely on trust. Whilst this may not be an issue for relatively simple calculations, it may be for complex algorithms, such as those involved with minimisation.

@@

\section*{Acknowledgements}
We acknowledge Anton Le Brun (ANSTO) for the provision of the lipid bilayer datasets in the example, James Hester (ANSTO) for comments made on the draft manuscript, and Andrew McCluskey (Bath University) and Isaac Gresham (UNSW) for important feedback on \refnx\ development.
